### 2021-05-28 submission/manuscript.Rmd
Finish draft plus edits

## Abstract
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C. difficile*-challenged mice. Administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc. When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90). Five of the top ten bacterial features important for predicting prolonged colonization had high relative abundances in the community. Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C. difficile* colonization in PEG-treated mice.

## Importance
Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are some microbiota similarities between diarrheal and *C. difficile* infection (CDI) communities such as lower diversity compared to healthy controls, which lead us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge Our findings suggest the osmotic laxative disrupts colonization resistance to *C. difficile*, as well as clearance in mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2014]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives in the context of CDIs. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C. difficile* clearance.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days: 1. received no additional treatment, 2. was also treated with clindamycin, and 3. was allowed to recover for 10 days prior to challenge (Fig. 1A). PEG treatment resulted in weight loss in all 3 groups of PEG-treated mice relative to their baseline weights, with the greatest change in weight observed on the fifth day of PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. However, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). In contrast, the clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). Therefore, PEG treatment led to sustained colonization in contrast to clindamycin mice that naturally cleared *C. difficile* within ten days post-challenge.

Notably, we also found PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumably due to *C. difficile* as the bacteria became detectable in stool samples from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

#List of variables that significantly explained variation in stool samples
f_d_top_effects <- f_d_sign_contrib %>%
  pull(effects)
#Sum of R2 values for variables that significantly explained variation in stool samples
f_d_top_R2_sum <- f_d_sign_contrib %>%
  select(R2) %>% sum
#P-value for variables that significantly explained variation in stool samples
f_d_top_p <- f_d_sign_contrib %>%
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text

@@

```
**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to coprophagy [@Nguyen2015]. We tried to minimize the impact of cage effects on our experiment by breaking up littermates when assigning mice to treatment groups. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and its interaction with treatment group did no significantly explain the observed variation in fecal communities (Data Set S1, Sheet X). None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, Sheet X). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, Sheet X). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

Interestingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine if there were genera that changed during the post-challenge period when the group median *C. difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig. S1A). We found no bacteria with relative abundances that were significantly different over the two time points after multiple hypothesis correction (Data Set S1, Sheet X). However, there was also wide variation between individual mice regarding when *C. difficile* became detectable (Fig. S1A) as well as the relative abundances of bacterial genera in the communities (Fig. S1B). For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period and this corresponded to mouse 10, which died on the sixth day post-challenge and mouse 11, where *C. difficile* was present at a high density from the 4th day post-challenge onward (Fig. S1B). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

#List of variables that significantly explained variation in tissue samples
f_d_top_effects_t <- f_d_sign_contrib_t %>%
  pull(effects)
#Sum of R2 values for variables that significantly explained variation in tissue samples
f_d_top_R2_sum_t <- f_d_sign_contrib_t %>%
  select(R2) %>% sum
#P-value for variables that significantly explained variation in tissue samples
f_d_top_p_t <- f_d_sign_contrib_t %>%
  select(`Pr(>F)`) %>% max # P value will be rounded in text

@@

```
**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig. 3A). Alpha diversity continued to increase over time based on the tissues from PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A, Data Set S1, Sheet X). Group, time point, and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)`) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set S1, Sheet X). Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C. difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge. This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Because there were differences in the mucosal microbiota including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon H&E stained histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5-2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

***C. difficile* challenge did not have a synergistic disruptive effect on the microbiota of PEG-treated mice.** Because *C. difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice. Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S3A-C). Similarly, there was no difference between the tissues of mock and *C. difficile* challenged mice (Fig. S3D-F). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.
``` {r 1-day PEG stool microbiota}

@@

```
**1-day laxative treatment resulted in transient *C. difficile* colonization and minor microbiota disruption.** Next, we examined how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C. difficile* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG treated mice were only transiently colonized and cleared *C. difficile* by 7 days post-challenge (Fig. 3B). The stool communities of PEG-treated mice were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig. 3C-D). We found the relative abundances of `r recovered_genera` genera were impacted by treatment, but recovered close to baseline levels by 7 days post-challenge including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, Sheet X). These findings suggest the duration of the PEG treatment was relevant, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

@@

```
**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild microbiota perturbation, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. To examine the impact of PEG treatment on *C. difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to our hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B).

>>>>

## Abstract
Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota. However, non-antibiotic medications such as the ubiquitous osmotic laxative polyethylene glycol (PEG) 3350, also alter the microbiota, but whether PEG impacts CDI susceptibility and clearance is unclear. To examine how PEG impacts susceptibility, we treated C57Bl/6 mice with 5-day and 1-day doses of 15% PEG in the drinking water and then challenged the mice with *C. difficile* 630. We used clindamycin-treated mice as a control because they consistently clear *C. difficile* within 10 days post-challenge (dpc). PEG treatment alone was sufficient to render mice susceptible and 5-day PEG-treated mice remain colonized for up to 30 dpc. In contrast, 1-day PEG treated mice were transiently colonized, clearing *C. difficile* within 7 dpc. To examine how PEG treatment impacts clearance, we administered a 1-day PEG treatment to clindamycin-treated, *C. difficile*-challenged mice. Administering PEG to mice after *C. difficile* challenge prolonged colonization up to 30 dpc. When we trained a random forest model with community data from 5 dpc, we were able to predict which mice would exhibit prolonged colonization (AUROC = 0.90). Examining the dynamics of these bacterial during the post-challenge period revealed patterns in the relative abundances of *Bacteroides*, *Enterobacteriaceae*, *Porphyromonadaceae*, *Lachnospiraceae*, and *Akkermansia* that were associated with prolonged *C. difficile* colonization in PEG-treated mice.

## Importance
Diarrheal samples induced by medications such as laxatives are typically rejected for *Clostridiodes difficile* testing. However, there are some microbiota similarities between diarrheal and *C. difficile* infection (CDI) communities such as lower diversity compared to samples from healthy patients, which lead us to hypothesize that diarrhea may be an indicator of *C. difficile* risk. We explored how osmotic laxatives disrupt the microbiota's colonization resistance to *C. difficile* by administering a laxative to mice either before or after *C. difficile* challenge. Our findings suggest the osmotic laxative disrupts colonization resistance to *C. difficile*, as well as clearance in mice already colonized with *C. difficile*. Considering that most hospitals recommend not performing *C. difficile* testing on patients taking laxatives and laxatives are used when administering fecal microbiota transplants via colonoscopy to patients with recurrent CDIs, further studies are needed to evaluate if laxatives impact microbiota colonization resistance in humans.

@@

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility [@Vandeputte2015]. Our group has shown that when *C. difficile* negative samples from patients are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and *C. difficile* negative patients with diarrhea compared to samples that were *C. difficile* negative with non-diarrheal stool consistency [@Schubert2014]. These results led to the hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs, regardless of how they developed diarrhea.

Osmotic laxatives can lead to diarrhea depending on the dose administered and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* colonization [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from CDIs by protecting intestinal epithelial cells from the toxins produced by *C. difficile* [@Olson2013; @Diebel2014]. Whether laxative administration results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

@@

Our group has used C57BL/6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [@Schubert2015; @Jenior2017; @Tomkovich2020]. Although two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile* [@Ferreyra2014; @VanInsberghe2020], these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance. Here, we characterized how long PEG-treated mice remain susceptible, whether PEG treatment results in more sustained *C. difficile* colonization and severe CDI than mice treated with clindamycin, and if PEG treatment after challenge can promote *C. difficile* clearance. Addressing these questions will better inform how we think about laxatives and diarrhea in the context of CDIs.

## Results
**5-day laxative treatment led to prolonged *C. difficile* colonization in mice.** Building off of previous work that showed treating mice with the osmotic laxative, PEG 3350, rendered mice susceptible to *C. difficile* colonization [@Ferreyra2014; @VanInsberghe2020], we decided to test how long *C. difficile* colonization is sustained and how long PEG-treated mice remain susceptible to *C. difficile*.  We compared three groups of mice treated with PEG 3350 to one group of mice treated with our standard 10 mg/kg clindamycin treatment, which temporarily renders the mice susceptible to *C. difficile* colonization, with mice typically clearing *C. difficile* within 10 days post-challenge [@Tomkovich2019; @Tomkovich2020]. All three groups of PEG-treated mice were administered a 15% PEG solution in the drinking water for 5-days. The first group received no additional treatment. The second group was also treated with clindamycin. A third group was allowed to recover for 10 days prior to challenge (Fig. 1A). The PEG treatment resulted in weight loss for the 3 groups of mice, with the greatest change in weight observed on the fifth day of the PEG treatment. The mice recovered most of the lost weight by five days after treatment (Fig. 1B). After either the PEG, clindamycin, or PEG and clindamycin treatment all mice were challenged with 10^5^ *C. difficile* 630 spores (Fig. 1A). All treatments rendered mice susceptible to *C. difficile* colonization. In contrast to the mice that only received clindamycin, PEG-treated mice remained colonized with *C. difficile* at a high level through thirty days post-challenge (Fig. 1C). The clindamycin-treated mice cleared *C. difficile* within ten days post-challenge (Fig. 1C). It was noteworthy that PEG-treated mice were still susceptible to *C. difficile* colonization after a 10-day recovery period, although *C. difficile* was not detectable in most of the group in the initial five days post-challenge (Fig. 1C, S1A). One mouse was found dead on the 6th day post-challenge, presumably due to *C. difficile* as the bacterium became detectable in stool samples from that mouse on the 4th day post-challenge (Fig. S1A, mouse 10). From 8 days post-challenge onward, the density of *C. difficile* stabilized in the 10-day recovery group and remained high through 20-30 days post-challenge (Fig. 1C). Thus, osmotic laxative treatment alone was sufficient to render mice susceptible to prolonged *C. difficile* colonization and PEG-treated mice remained susceptible through ten days post-treatment.

@@

#List of variables that significantly explained variation in stool samples
f_d_top_effects <- f_d_sign_contrib %>%
  pull(effects)
#R2 for time (day)
r2_day <- f_d_sign_contrib %>% 
  filter(effects == "day") %>% 
  pull(R2)
#R2 for group
r2_group <- f_d_sign_contrib %>% 
  filter(effects == "group") %>% 
  pull(R2)
#Sum of R2 values for variables that significantly explained variation in stool samples
f_d_top_R2_sum <- f_d_sign_contrib %>%
  select(R2) %>% sum
#P-value for variables that significantly explained variation in stool samples
f_d_top_p <- f_d_sign_contrib %>%
  select(`Pr(>F)`) %>% max # Both p values are the same, will be rounded in text

@@

```
**5-day laxative treatment differentially disrupted the fecal microbiota compared to clindamycin treatment.** Since osmotic laxatives and clindamycin have previously been shown to disrupt the murine microbiota [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020], we hypothesized the different *C. difficile* colonization dynamics between mice treated with the osmotic laxative or clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time (R^2^ = `r r2_day`) and treatment group (R^2^ = `r r2_group`) explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage, and sequencing preparation plate (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, sheet 1). None of the treatment groups recovered to their baseline community structure either 10 or 30 days post-challenge suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Since none of the communities completely recovered in the follow-up period after treatments, we next profiled community diversity and composition. We examined the alpha diversity dynamics by calculating the communities' Shannon diversity. Although both clindamycin and PEG treatments decreased diversity, Shannon diversity was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-challenge (Fig. 2C; Data Set S1, sheet 2). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-PEG-treatment. We found `r PEG_genera` genera that were altered by PEG treatment (Data Set S1, sheet 3). The majority of the bacterial relative abundances decreased after the PEG treatment, but the relative abundance among members of the *Enterobacteriaceae* and *Bacteroides* increased. The increase in *Bacteroides* relative abundance was unique to PEG treated mice, as the *Bacteroides* relative abundance actually decreased in clindamycin treated mice (Fig. 2D). Finally, we identified the genera whose relative abundance differed across treatment groups over multiple time points. Of the `r f_d_group_total_genera` genera that were different between treatment groups, `r f_d_group_multi_day` genera were different over multiple time points  (Fig. 2E, Data Set S1, sheet 4). Thus, PEG had a significant impact on the fecal microbiota that was maintained over time and was distinct from clindamycin treatment.

Because *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge, we decided to examine if there were genera that changed during the post-challenge period. We compared the communities from when *C. difficile* shifted from undetectable at 1 day post-challenge to detectable in the stool samples with the density stabilizing around 8 days post-challenge (Fig. S1A). We found no genera with relative abundances that were significantly different over the two time points (Data Set S1, sheet 5). However, there was also wide variation between individual mice regarding when *C. difficile* became detectable (Fig. S1A) as well as the relative abundances of bacterial genera in the communities (Fig. S1B). For example, two mice had a high relative abundance of *Enterobacteriaceae* throughout the post-challenge period. One mouse died on the sixth day post-challenge and in the other *C. difficile* was present at a high density from the 4th day post-challenge onward (Fig. S1B). While we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay was striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

@@

#List of variables that significantly explained variation in tissue samples
f_d_top_effects_t <- f_d_sign_contrib_t %>%
  pull(effects)
#R2 for time (day)
r2_day_t <- f_d_sign_contrib_t %>% 
  filter(effects == "day") %>% 
  pull(R2)
#R2 for group
r2_group_t <- f_d_sign_contrib_t %>% 
  filter(effects == "group") %>% 
  pull(R2)
#Sum of R2 values for variables that significantly explained variation in tissue samples
f_d_top_R2_sum_t <- f_d_sign_contrib_t %>%
  select(R2) %>% sum
#P-value for variables that significantly explained variation in tissue samples
f_d_top_p_t <- f_d_sign_contrib_t %>%
  select(`Pr(>F)`) %>% max # P value will be rounded in text

@@

```
**5-day laxative treatment did not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced communities associated with tissues collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, the alpha diversity was lower in the PEG-treated mice compared to clindamycin treated mice (Fig. 3A, Data Set S1, sheet 6). The alpha diversity of the tissue-associated community increased in PEG-treated mice collected at 20 and 30 days post-challenge (Fig. 3A). Group (R^2^ = `r r2_group_t`), time point (R^2^ = `r r2_day_t`), and their interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, sheet 7). We saw the greatest difference in the relative abundance of the mucosal microbiota between treatment groups (clindamycin, 5-day PEG, and 5-day PEG plus clindamycin) at 6 days post-challenge with `r shared_tissue_genera` genera that were significantly different (*P* `r format.pval(tissue_genera_max_p, eps = 0.05)`) in all three of the tissue types we collected (cecum, proximal colon, and distal colon; Fig. S2A, Data Set S1, sheet 8, 9, and 10). Interestingly, *Peptostreptococcaceae* (the family with a sequence that matches *C. difficile*) was one of the genera that had a significant difference in relative abundance between treatment groups at 6 days post-challenge. This population was primarily only present in the 5-day PEG treatment group of mice and decreased in the proximal and distal colon tissues over time (Fig. S2B). By 30 days post-challenge, only the relative abundances of *Bacteroides*, *Clostridiales*, *Firmicutes*, and *Ruminococcaceae* were different between treatment groups and only in the cecum tissues (Fig. 3C, Fig. 2E, Data Set S1, sheet 8). Thus, PEG treatment had a significant impact on the mucosal microbiota and we detected *C. difficile* sequences in the cecum, proximal colon, and distal colon tissue communities.

Because there were differences in the mucosal microbiota including detectable *C. difficile* sequences in tissues from PEG-treated mice relative to mice treated with clindamycin, we next examined the severity of *C. difficile* challenge by evaluating cecum and colon histopathology [@Reeves2011]. However, we found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-challenge (Fig. 3D), the time point typically examined in *C. difficile* 630 challenged mice [@Dieterle2020]. We also looked at 6 days post-challenge because that was when there was a large difference in *C. difficile* density between PEG- and clindamycin-treated mice (Fig. 1C). Although there was a slight difference in the histopathology score of the colon between PEG and clindamycin-treated mice, there was not a signifant difference in the cecum and the overall score was relatively low (1.5 to 2.5 out of 12, Fig. 3E). Therefore, although PEG treatment had a disruptive effect on the mucosal microbiota, the impact of *C. difficile* 630 challenge on the cecum and colon was similar between PEG and clindamycin treated mice.

***C. difficile* challenge did not have a synergistic disruptive effect on the microbiota of PEG-treated mice.** Because *C. difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged mice treated with clindamycin or PEG. Examining the stools of the mock-challenged mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S3A-C). Similarly, there was no difference between the communities found in the tissues of mock and *C. difficile* challenged mice (Fig. S3D-F). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.
``` {r 1-day PEG stool microbiota}

@@

```
**1-day laxative treatment resulted in transient *C. difficile* colonization and minor microbiota disruption.** Next, we examined how a shorter osmotic laxative perturbation would impact the microbiome and susceptibility to *C. difficile*. We administered either a 1-day PEG treatment, a 1-day PEG treatment with a 1-day recovery period, or clindamycin to mice before challenging them with *C. difficile* (Fig. 3A). In contrast to the 5-day PEG treated mice, the 1-day PEG groups were only transiently colonized and cleared *C. difficile* by 7 days post-challenge (Fig. 3B). The stool communities of the 1-day PEG treatment groups were also only transiently disrupted, with Shannon diversity recovering by 7 days post-challenge (Fig. 3C-D, Data Set S1, sheets 11 and 12). We found the relative abundances of `r recovered_genera` genera were impacted by treatment, but recovered close to baseline levels by 7 days post-challenge including *Enterobacteriaceae*, *Clostridiales*, *Porpyromonadaceae*, and *Ruminococcaceae* (Fig. 3E, Data Set S1, sheet 13 and 14). These findings suggest the duration of the PEG treatment was relevant, with shorter treatments resulting in a transient loss of *C. difficile* colonization resistance.

@@

```
**Post-challenge laxative treatment disrupted clearance in clindamycin-treated mice regardless of whether an FMT was also administered.** Since a 1-day PEG treatment resulted in a more mild perturbation of the microbiota, we decided to use the 1-day treatment to examine the hypothesis that PEG helps to flush *C. difficile* spores from the intestine. This hypothesis is proposed in the discussion section of FMT studies where bowel prep is part of the preparation undergone by patients receiving FMTs via colonoscopy [@vanNood2013; @Razik2017; @Postigo2012; @Liacouras1996]. To examine the impact of PEG treatment on *C. difficile* clearance, we treated 4 groups of mice with clindamycin and then challenged all mice with *C. difficile* before administering the following treatments: no additional treatment, 1-day PEG immediately after challenge, and 1-day PEG treatment 3 days after challenge followed by either administration of an FMT or PBS solution by oral gavage (Fig. 5A). Contrary to the hypothesis, all groups of mice that received PEG exhibited prolonged *C. difficile* colonization (Fig. 5B).
